Fast Bayesian Inference for Exoplanet Discovery in Radial Velocity Data

### 2015-01-27 Paper/ms.tex
A bit of reordering

\begin{abstract}
Inferring the number of planets $N$ in an exoplanetary system from radial velocity
(RV) data is a challenging task. Recently, it has become clear that RV data
can contain periodic signals due to stellar activity, which can be difficult
to distinguish from planetary signals. However, even doing the inference
under a given set of simplifying assumptions (e.g. no stellar activity) can
be difficult. It is common for the posterior distribution for the
planet parameters, such as orbital periods,
to be multimodal and to have other awkward features. In
addition, when $N$ is unknown, the marginal likelihood (or evidence) as a
function of $N$ is required. Rather than doing separate runs with different
trial values of $N$, we propose an alternative
approach using a trans-dimensional Markov Chain Monte Carlo method within
Nested Sampling. The posterior distribution for $N$ can be obtained with a
single run. We apply the method to $\nu$ Oph and Gliese 581, finding moderate evidence
for additional signals in
$\nu$ Oph with periods of 36.11 $\pm$ 0.034 days,
75.58 $\pm$ 0.80 days, and 1709 $\pm$ 183 days;
the posterior probability that at least one of these
exists is 85\%. The results also suggest Gliese 581 hosts many (7-15) ``planets''
(or other causes of other periodic
signals), but only four have well determined periods. The analysis of both
of these datasets shows phase transitions exist which are difficult to
negotiate without Nested Sampling.
\end{abstract}

@@

It is now recognised that
there are many possible sources for oscillations in a data set and not all
such oscillations should be claimed as planets. 
Our model cannot distinguish between oscillations due to planets and
oscillations due to stellar activity: any oscillations found in the dataset
will be described as ``planets'' by the model. Another important consideration
is the physical stability of the orbital system which is ignored by this type
of analysis \citep{orbital}. However it is interesting that
we find many more signals in the data than previous authors.
By inspecting the posterior distribution for the periods
(Figure~\ref{fig:gliese581_periods}), we see that only four of the periods
are well determined, corresponding to the known periods of Gliese 581 b, c, d,
and e. The other ``periods'' are very uncertain. It is quite likely that a
non-sinusoidal signal due to stellar activity is being modelled as several
periods \citep{astero}, and if the model were extended to include a ``stochastic''
oscillation \citep{gaussproc}, the number of periods detected may be reduced
substantially. With these kinds of models, the posterior distribution for $N$
can also be influenced by the prior for the amplitudes $A_i$. Many authors
assign independent broad priors to the amplitudes, and this causes the
``Occam's razor'' penalty for adding extra signals to be quite strong. Since we
use a hierarchical prior for the amplitudes, if some amplitudes are found to
be low, $\mu_A$ will become small. When $\mu_A$ is small, it is likely that any
extra signals will have small amplitudes, so the ``Occam's razor'' effect is
weaker. An example model fit for Gliese 581 is shown in
Figure~\ref{fig:gliese581}.

@@

As with $\nu$ Oph, we can calculate the posterior probability of any hypothesis
about Gliese 581 by computing the fraction of the posterior samples that have
that property. The posterior probabilities for planets b, c, d, and e, are
close to 1. The posterior probability a signal exists with $\log_{10}(P)
\in [1.55, 1.57]$ is 88\%, and the probability for a signal with
$\log_{10}(P) \in [2.6, 2.8]$ is 85\%.

>>>>

\begin{abstract}
Inferring the number of planets $N$ in an exoplanetary system from radial velocity
(RV) data is a challenging task. Recently, it has become clear that RV data
can contain periodic signals due to stellar activity, which can be difficult
to distinguish from planetary signals. However, even doing the inference
under a given set of simplifying assumptions (e.g. no stellar activity) can
be difficult. It is common for the posterior distribution for the
planet parameters, such as orbital periods,
to be multimodal and to have other awkward features. In
addition, when $N$ is unknown, the marginal likelihood (or evidence) as a
function of $N$ is required. Rather than doing separate runs with different
trial values of $N$, we propose an alternative
approach using a trans-dimensional Markov Chain Monte Carlo method within
Nested Sampling. The posterior distribution for $N$ can be obtained with a
single run. We apply the method to $\nu$ Oph and Gliese 581, finding moderate evidence
for additional signals in
$\nu$ Oph with periods of 36.11 $\pm$ 0.034 days,
75.58 $\pm$ 0.80 days, and 1709 $\pm$ 183 days;
the posterior probability that at least one of these
exists is 85\%. The results also suggest Gliese 581 hosts many (7-15) ``planets''
(or other causes of other periodic
signals), but only 4-6 have well determined periods. The analysis of both
of these datasets shows phase transitions exist which are difficult to
negotiate without Nested Sampling.
\end{abstract}

@@

It is now recognised that
there are many possible sources for oscillations in a data set and not all
such oscillations should be claimed as planets. 
Our model cannot distinguish between oscillations due to planets and
oscillations due to stellar activity: any oscillations found in the dataset
will be described as ``planets'' by the model. Another important consideration
is the physical stability of the orbital system which is ignored by this type
of analysis \citep{orbital}. However it is interesting that
we find many more signals in the data than previous authors.
By inspecting the posterior distribution for the periods
(Figure~\ref{fig:gliese581_periods}), we see that only four of the periods
are well determined, corresponding to the known periods of Gliese 581 b, c, d,
and e. The other ``periods'' are more uncertain. As with $\nu$ Oph, we can
calculate the posterior probability of any hypothesis
about Gliese 581 by computing the fraction of the posterior samples that have
that property. The posterior probabilities for planets b, c, d, and e, are
close to 1. The posterior probability a signal exists with $\log_{10}(P)
\in [1.55, 1.57]$ is 88\%, and the probability for a signal with
$\log_{10}(P) \in [2.6, 2.8]$ is 85\%.

One possible explanation for the large number of inferred signals that a
non-sinusoidal signal due to stellar activity is being modelled as several
periods \citep{astero}, and if the model were extended to include a ``stochastic''
oscillation \citep{gaussproc}, the number of periods detected may be reduced
substantially. With these kinds of models, the posterior distribution for $N$
can also be influenced by the prior for the amplitudes $A_i$. Many authors
assign independent broad priors to the amplitudes, and this causes the
``Occam's razor'' penalty for adding extra signals to be quite strong. Since we
use a hierarchical prior for the amplitudes, if some amplitudes are found to
be low, $\mu_A$ will become small. When $\mu_A$ is small, it is likely that any
extra signals will have small amplitudes, so the ``Occam's razor'' effect is
weaker. An example model fit for Gliese 581 is shown in
Figure~\ref{fig:gliese581}.

@@
